We are not the 99 percent: quantifying asphericity in the distribution of Local Group satellites

### 2018-02-21 paper/paper.tex
references + intro

Later, \cite{2007MNRAS.374.1125M} used simulations to confirm the low
chances ($<0.5\%$) found in \cite{2005A&A...431..517K}.
This result was challenged by \cite{2009MNRAS.399..550L} who continue
to report high chances to find a planar configuration;
two more recent numerical experiment with high resolution simulations
(\cite{2013MNRAS.429..725S} and \cite{2016MNRAS.457.1931S}) report
contradicting results (low and high chances to find the observational
result, respectively) but precise figures are not quoted in either of
these three reports. \citep{2015ApJ...815...19P,2016MNRAS.457.1931S}.
%\citep{2015ApJ...815...19P,2016MNRAS.457.1931S}.
The presence of a Vast Polar Structure of satellites around the Milky
Way, stablished in the last decade \citep{2013MNRAS.435.1928P}.
This structure is present in the $11$ brightest "classical" dwarf
galaxies and has been confirmed as fainter satellites have been
detected \citep{2014MNRAS.441.2124B}. 
In the case of M31, observational studies have found a planar satellite
distribution in a subset of $15$ satellites out of the 
total poulation of $27$ satellites.\citep{2013ApJ...766..120C,
2013Natur.493...62I}, although its brightest satellites
do not show any special feature and are consistent with a random
spherically symmetric distribution \citep{2006AJ....131.1405K,
  2007MNRAS.374.1125M, 2013ApJ...766..120C}.

The question whether this kind of planar configurations are atypical
on cosmological samples has not been properly addressed given the
current observational limitations. For instance, comparisons in large
galaxy surveys have only been able to constrain that $5\%$ of the
Milky-Way like galaxies in the Universe have two bright satellite
galaxies as is the case of the Milky Way and its Large and Small
Magellanic Clouds \citep{2011ApJ...733...62L}, or
that pairs of bright satellites preferentially show anti-correlated
when they are diametrically oposed \citep{2014Natur.511..563I}.
A planar structure outside the Local Group (LG) has been confirmed
in Centaurus A \citep{muller2018whirling} a galaxy that is only $\sim
4$Mpc away.

To date, the atipicallity of satellite distributions can only be
addressed in simulations.
First studies 
Most of the studies have been performed in the currently preferred
paradigm, which is based on a Cold Dark Matter cosmology in a expanding
Universe described by General Relativity including a Cosmological
Constant, the so-called Lambda Cold Dark Matter (LCDM) model.
The broad consensus is that these highly structured satellite distributions
are hard to find in simulations. 
There are still disagreements on the precise degree of atypicallity. 
\cite{2015MNRAS.452.3838C} showed that part of the disagreement is the
ambiguitiy in defining a planar configuration, finding that actually
planar structures show a large diversity, hindering the possibility to
use one or to systems as a cosmological probe.

@@

\begin{table*}
\centering
\begin{tabular}{|p{4.0cm}|p{5.5cm}| p{4.0cm}| c|}\hline
Reference & Target Measurement & Parent Sample & Probability ($\%$)\\\hline
\text{\cite{2005A&A...431..517K}} & RMS height in MW's 11 classical
satellites & Monte Carlo from an spherical power law
radial distribution. & $<0.5$ \\
\text{\cite{2005MNRAS.363..146L}} & Triaxiality of MW's 11 classical
satellites & 6 high resolution DM only N-body simulations ($\sim10^5$
\Msun\ particle mass resolution). & High\\ 
\text{\cite{2005ApJ...629..219Z}} & Triaxiality and disk height in 
MW's 11 classical satellites & 3 high resolution DM only N-body
simulations. ($\sim10^6$ \Msun\ particle mass resolution) & 2 \\
\text{\cite{2007MNRAS.374.1125M}} & Triaxiality and RMS height in MW's
11-13 brightest satellites & Monte Carlo from a halo triaxiality distribution from LCDM
simulations & $<0.5$\\
\text{\cite{2009MNRAS.399..550L}}& Triaxiality in MW 11 classical satellites & 436 halos from a
cosmological N-body simulation ($1.3\times 10^{8}$\Msun\ particle mass)
&  High \\
\text{\cite{2013MNRAS.429..725S}}& $c/a$ ratio in MW's 12 brightest
satellites & 6 High resolution DM only N-body simulations ($\sim
10^3$\Msun particle mass resolution) of individual halos & Low \\
\text{\cite{2014ApJ...789L..24P}}& $c/a$ and $b/a$ ratio in MW's 11
classical satellites & 48 high resoltion DM only N-body simulations
($\sim 10^{5}$\Msun particle mass resolution) of both halo pairs and
isolated halos & 0.77\\
\text{\cite{2016MNRAS.457.1931S}}& $c/a$ ratio in MW's 11 classical satellites & 12
high resolution Hydro simulation of halo pairs ($\sim 10^{4}$\Msun particle
mass resolution) & High\\
This Work & Triaxiality and RMS height in M31's 11-15 brightest
satellites & 24 halos from selected pairs in a cosmological N-body
Dark Matter only simulation ($\sim 10^{6}$\Msun particle mass resolution)& 3\\
This Work & Triaxiality and RMS height in M31's 11-15 brightest
satellites & 20 halos from selected pairs in acosmological N-body
hydro simulation ($\sim 10^{6}$\Msun particle mass resolution)& 0.6 \\
This Work & Triaxiality and RMS height in M31's 11-15 brightest
satellites & 12 high resolution DM only N-body simulations ($\sim
10^{5}$\Msun particle mass resolution) of halo pairs & 0.2 \\
\hline
\end{tabular}
\caption{Same as Table \ref{table:M31} for the MW satellites.
\label{table:MW}}
\end{table*}

Some studies have focused on the individual dark matter halos
that could host galaxies like the MW or M31, while some other studies
have tried to simulate the formation of galaxy pairs resembling both
the MW and M31.
The transition from individual halos to pairs has been
motivated by studies that show how the LG location in the cosmic web
plays a role in distinguishing the MW and M31 halos from the
average $\sim 10^{12}$ \Msun\ DM halo \citep{2011MNRAS.417.1434F}
Studies of the LG as a pair of isolated DM halos have also found that
the LG itself has relatively atypical kinematics in the context
of LCDM and a strong preference to lie along filaments \citep{ForeroRomero2013}.
Putting together these pieces, the cosmic web should therefore play a
determinant role in building up planar satellite distributions through
preferential alignments and accretions histories of galaxies
\citep{2014MNRAS.443.1090F}.

>>>>

Later, \cite{2007MNRAS.374.1125M} used simulations to confirm the low
chances ($<0.5\%$) found in \cite{2005A&A...431..517K}.
This result was challenged by \cite{2009MNRAS.399..550L} who continue
to report high chances to find a planar configuration;
two more recent numerical experiment with high resolution simulations
(\cite{2013MNRAS.429..725S} and \cite{2016MNRAS.457.1931S}) report
contradicting results (low and high chances to find the observational
result, respectively) but precise figures are not quoted in either of
these three reports. \citep{2015ApJ...815...19P,2016MNRAS.457.1931S}.
Meanwhile, \cite{2013MNRAS.429.1502W, 2014ApJ...789L..24P} reported
on new tests using high resolution simulations and cosmological
volumes arriving at a chance of $6\%$ and $0.77\%$, respectively to
have a satellite distribution as triaxial as the observations.

In the case of M31, observational studies have found a planar satellite
distribution in a subset of $15$ satellites out of the 
total poulation of $27$ satellites \citep{2013ApJ...766..120C,
2013Natur.493...62I}, although its brightest satellites
do not show any special feature.
All the studies agree on the point that the satellite distribution of
the 15 to 27 brightest satellites are consistent with 
spherically symmetric distribution and easy to reproduce in LCDM
simulations \citep{2006AJ....131.1405K,
  2007MNRAS.374.1125M, 2013ApJ...766..120C}.

@@

\begin{table*}
\centering
\begin{tabular}{|p{4.0cm}|p{4.5cm}| p{5.5cm}| c|}\hline
Reference & Target Measurement & Parent Sample & Probability ($\%$)\\\hline
\text{\cite{2005A&A...431..517K}} & RMS height in 11 classical
satellites & Monte Carlo from an spherical power law
radial distribution. & $<0.5$ \\
\text{\cite{2005MNRAS.363..146L}} & Triaxiality of 11 classical
satellites & 6 high resolution DM only N-body simulations ($\sim10^5$
\Msun\ particle mass resolution). & High\\ 
\text{\cite{2005ApJ...629..219Z}} & Triaxiality and disk height in 
11 classical satellites & 3 high resolution DM only N-body
simulations. ($\sim10^6$ \Msun\ particle mass resolution) & 2 \\
\text{\cite{2007MNRAS.374.1125M}} & Triaxiality and RMS height in 11-13 brightest satellites & Monte Carlo from a halo triaxiality distribution from LCDM
simulations & $<0.5$\\
\text{\cite{2009MNRAS.399..550L}}& Triaxiality in 11 classical satellites & 436 halos from a
cosmological N-body simulation ($1.3\times 10^{8}$\Msun\ particle mass)
&  High \\
\text{\cite{2013MNRAS.429..725S}}& $c/a$ ratio in 12 brightest
satellites & 6 High resolution DM only N-body simulations ($\sim
10^3$\Msun particle mass resolution) of individual halos & Low \\
\text{\cite{2013MNRAS.429.1502W}}& $c/a$ ratio or RMS height in 11 brightest
satellites & 1686 halos from a cosmological DM only N-body simulation
($\sim 10^6$\Msun particle mass resolution) & 6 or 13 \\
\text{\cite{2014ApJ...789L..24P}}& $c/a$ and $b/a$ ratio in 11
classical satellites & 48 high resoltion DM only N-body simulations
($\sim 10^{5}$\Msun particle mass resolution) of both halo pairs and
isolated halos & 0.77\\
\text{\cite{2016MNRAS.457.1931S}}& $c/a$ ratio in 11 classical satellites & 12
high resolution Hydro simulation of halo pairs ($\sim 10^{4}$\Msun particle
mass resolution) & High\\
This Work & Triaxiality and RMS height in 11-15 brightest
satellites & 24 halos from selected pairs in a cosmological N-body
Dark Matter only simulation ($\sim 10^{6}$\Msun particle mass resolution)& 3\\
This Work & Triaxiality and RMS height in 11-15 brightest
satellites & 20 halos from selected pairs in acosmological N-body
hydro simulation ($\sim 10^{6}$\Msun particle mass resolution)& 0.6 \\
This Work & Triaxiality and RMS height in 11-15 brightest
satellites & 12 high resolution DM only N-body simulations ($\sim
10^{5}$\Msun particle mass resolution) of halo pairs & 0.2 \\
\hline
\end{tabular}
\caption{Same as Table \ref{table:M31} for the MW satellites.
\label{table:MW}}
\end{table*}
